Generalized Quasi-linear Approximation and Non-normality in Taylor-Couette Spiral Turbulence

### 2022-08-15 paper/paper.tex
Updates from Overleaf

In order to demonstrate the effects non-normality plays, figure~\ref{fig:pseudospectra} shows both the modal spectrum (in blue points) and the $\epsilon-$pseudospectrum for the supercritical and subcritical $\Reyn_i$. The contour value $\epsilon^{-1}$ represents the set of values 
\begin{equation}
\label{eqn:pseudospectrum}
    \lVert (\sigma \mathbf{I} - \Lop)^{-1} \rVert \geq \epsilon^{-1}
\end{equation}
for a given complex number $\sigma = \gamma + i\omega$ here representing the eigenvalue corresponding to the temporal evolution of a linear solution, $e^{\sigma t}$ (see \cite{trefethen2005spectra} for an introduction). Equation~\ref{eqn:pseudospectrum} depends on the choice of norm; here we use the standard energy norm. These contours in turn give a measure of the maximum amplification of a linear disturbance: the further negative contours extend into the upper-half plane, the larger a linear disturbance can grow as a result of non-normality, irrespective of the presence of instability. This linear effect is captured entirely by GQL (and QL, for that matter). What role it plays in the development of subcritical spiral turbulence remains an open question.
\begin{figure}
    \centering
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_900.00_eta_0.8830_mu_-3.3333_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_700.00_eta_0.8830_mu_-4.2857_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \caption{Spectra (blue points) and pseudospectra (contours) for $\Reyn_i = 900$ (left) and $\Reyn_i = 700$ (right) for a mode with $m=6$ and $k_z = 9.68$. The latter is the z wavenumber at the critical $\Reyn_{i,crit} \simeq 778$. $\omega$ is the frequency and $\gamma$ the growth rate at each point in the domain. The stability boundary is marked with a gray line at $\gamma = 0$. The are seven equally spaced contours from $-1.8$ to 0.}
    \label{fig:pseudospectra}
\end{figure}
\section{Results}
\label{sec:nonlinear}
Figure~\ref{fig:urms_tz_rei700} shows a typical DNS snapshot of the rms velocity perturbation $u'_{rms}$ in the middle of the gap for $\Reyn_i=700$. Despite being well below the linear stability threshold $\Reyn_{i,crit} \simeq 778$, the flow morphology is nearly identical to the supercritical case $\Reyn_i = 900$ (not shown). 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{../figs/urms_tz_rei_700_reo-3398_000278.png}
    \caption{Slice of fluctuating rms velocity in the middle of the radial domain at $\Reyn_i = 700$, $\Reyn_o = -3398$. This solution is well below the linear stability threshold at }
    \label{fig:urms_tz_rei700}
\end{figure}
\subsection{Supercritical Spiral Turbulence}
We first consider the supercritical case with $\Reyn_i = 900$. We initialize GQL simulations with $\Lambda = 0, 1, 3, 5, 10$ using a snapshot of DNS data at $t \simeq 1.5 \tau_\nu$.  Figure~\ref{fig:rei900_snapshots} shows slices at mid-gap, $r = R_{mid}$. The first important thing to note is that none of these low-order models are able to maintain the spiral structure. However, even including a single additional low mode, $\Lambda = 1$, shows a significant improvement in flow morphology compared to QL. It is quite interesting to note that QL \emph{does} feature some degree of spatial intermittency--there are clearly separated regions hosting spiral waves interspersed with featureless, laminar patches. 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figs/rei900_snapshots.pdf}
    \caption{RMS velocity perturbations in the $z-\theta$ plane at $r=R_{mid}$ for $\Reyn_i=900, \Reyn_0=-3398$. The upper left panel shows DNS, the upper right panel shows QL, and the remaining four are GQL with $\Lambda = 1,3,5,10$. Even at $\Lambda = 10$, spiral turbulence is not apparent.}
    \label{fig:rei900_snapshots}
\end{figure}

Figure~\ref{fig:ke_vs_t_rei900} shows the kinetic energy as a function of time and $\Lambda$ for both the supercritical (left) and subcritical (right) runs. 
The supercritical case shows significant oscillations at all $\Lambda$. 
\begin{figure}
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/rei900_reo_-3398_KE_vs_t.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/rei700_reo_-3398_KE_vs_t.pdf}
    \end{minipage}
    \caption{Perturbation kinetic energy as a function of time for $\Reyn_o=-3398$. $\Reyn_i = 900$ (left), started from small perturbations, is above the linear stability boundary. $\Reyn_i = 700$ (right panel), continued by lower $\Reyn_i$ from previous simulations, is in the hysteretic region below the stability boundary. Starting from a saturated spiral turbulence state from DNS at $t/\tau_\nu \simeq 1.5$ (left), $t/\tau_\nu \simeq 3$ (right), we continue the solution with QL ($\Lambda = 0$) and GQL with cutoffs $\Lambda = 1,3,5,10,30,60$. At higher $\Reyn_i$, all of the reduced models begin high amplitude oscillations. At lower $\Reyn_i$, $\Lambda = 60$ is required to maintain the spiral turbulence state. }
    \label{fig:ke_vs_t_rei900}
\end{figure}
Focusing on $\Lambda=5$, figure~\ref{fig:rei900_lambda5_story} shows that the flow goes through a series of different states bearing a strong resemblance to other well-known TC flow patterns originally identified in exhaustive series of experiments reported in \cite{1984JFM...146...45M} and subsequently found in simulations (e.g. \cite{2009PhRvE..80d6315M}). At the time of kinetic energy minimum, labelled 1 in the left hand side of figure~\ref{fig:rei900_lambda5_story}, the flow seems to recapitulate the interpenetrating spirals (IPS) characteristic of lower $\Reyn$ counter-rotating. As the kinetic energy rises, reaching a knee at point 2, the flow morphology resembles the patchy bursts characteristic of the intermittency regime (INT). Finally, at the peak of the kinetic energy oscillation, point 3 in figure~\ref{fig:rei900_lambda5_story}, the flow reaches is closests approach to something resembling spiral turbulence, though it is very difficult to identify the flow as such from a slice at a single point in time and plane in space.

This is notable for a few reasons. At $\Reyn_i=900, \Reyn_o=-3398$ the system is not in a hysteretic region of parameter space. Thus, we do not expect this behaviour to be due to GQL picking up other, coexisting solutions. 
It also highlights the delicate balance an intermittent laminar/turbulent solution represents. The turbulence in this solution has a very broad range of scales, as evidenced by figure~\ref{fig:urms_tz_rei700}. Understanding how it exchanges energy with the mean flow is crucial to understanding both the saturation of the underlying linear instability as well as the maintenance of the non-linear state. 
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/rei900_story.pdf}
    \caption{Kinetic energy (left) and flow morphologies (right) for $\Reyn_i = 900$, $\Lambda = 5$ covering one full oscillation cycle. The panels on the right show three times corresponding to the vertical lines in the kinetic energy plot on the left. At the kinetic energy minimum (point 1), the flow shows a IPS-like morphology, while an INT burst occurs at the knee (point 2), and a more space-filling turbulent state occurs at maximum energy (point 3).}
    \label{fig:rei900_lambda5_story}
\end{figure}
Finally, in figure~\ref{fig:rei900_vmean}, we show the azimuthal velocity averaged over $\theta$ and $z$, giving the mean deviation from CCF as a function of GQL cutoff $\Lambda$. Note that no conditional averaging \cite{2009PhRvE..80f7301D} is performed here, so this represents an average over both laminar and turbulent regions for DNS. As $\Lambda$ is increased, the profile approaches the DNS case; this is typical of GQL behaviour (figure 7 of \cite{2017JFM...810..412T} shows the same for rotating plane Couette flow). However, unusually, QL performs slightly better than low order GQL with $\Lambda < 5$. 
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/rei900_reo_-3398_vmean_profile.pdf}
    \caption{Mean deviation from CCF for $\Reyn_i = 900$, $\Reyn_o=-3398$. The heavy line is DNS, the orange line is QL, and all others are GQL with $\Lambda = 1,3,5,10$.}
    \label{fig:rei900_vmean}
\end{figure}

@@

\section{Conclusion}
\label{sec:conclusion}
We have presented a series of simulations demonstrating the performance of GQL at various cutoffs for Taylor-Couette flow in the spiral turbulence regime. GQL is a significant improvement over QL models in representing mean quantities, including the deviation from circular Couette flow. However, even at the highest cutoffs studied here, it does not retain an unambiguous spiral form in either the sub- or supercritical regimes. Prior work has focused on the utility of GQL in presenting a more accurate \emph{statistical} picture of turbulent flows than QL \cite{2016PhRvL.116u4501M, 2017JFM...810..412T,2018RSPSA.47480422T,2019Kellam}. This focus is particularly important in light of its potential as as means of improving the accuracy of direct statistical simulation (DSS), a method for studying turbulence by computing its statistics directly, rather than individual realizations \cite{2022arXiv220505513M}.

However, we have shown here that it is possible to use GQL as a means to investigate the non-linear interactions of greatest importance for maintaining the intricate patterns manifested in Taylor-Couette flow. In future work, we will consider the ability of GQL to maintain exact coherent states identified by Deguchi et al \cite{2014PhRvL.112r4502D} and the dynamics of which were studied recently by Wang and coworkers \cite{2022arXiv220712990W}. Understanding how these interactions work to shape subcritical TC turbulence represents a promising new avenue to understanding how ECS are maintained and how they in turn sustain turbulence below linear stability bounds. In particular, one can begin to understand the relative importance of non-normal effects (represented entirely with GQL) and non-linear ECS.

>>>>

In order to demonstrate the effects non-normality plays, figure~\ref{fig:pseudospectra} shows both the modal spectrum (in blue points) and the $\epsilon-$pseudospectrum for the supercritical and subcritical $\Reyn_i$. The contour value $\epsilon^{-1}$ represents the set of values 
\begin{equation}
\label{eqn:pseudospectrum}
    \lVert (\sigma \mathbf{I} - \Lop)^{-1} \rVert \geq \epsilon^{-1}
\end{equation}
for a given complex number $\sigma = \gamma + i\omega$ here representing the eigenvalue corresponding to the temporal evolution of a linear solution, $e^{\sigma t}$ (see \cite{trefethen2005spectra} for an introduction). Equation~\ref{eqn:pseudospectrum} depends on the choice of norm; here we use the standard energy norm. These contours in turn give a measure of the maximum amplification of a linear disturbance: the further negative contours extend into the upper-half plane, the larger a linear disturbance can grow as a result of non-normality, irrespective of the presence of instability. This linear effect is captured entirely by GQL (and QL, for that matter). What role it plays in the development of subcritical spiral turbulence remains an open question.
\begin{figure}
    \centering
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_900.00_eta_0.8830_mu_-3.3333_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/TC_pseudospectrum_Re1_700.00_eta_0.8830_mu_-4.2857_nr_64_m_6_kz_ 9.68.pdf}
    \end{minipage}
    \caption{Spectra (blue points) and pseudospectra (contours) for $\Reyn_i = 900$ (left) and $\Reyn_i = 700$ (right) for a mode with $m=6$ and $k_z = 9.68$. The latter is the z wavenumber at the critical $\Reyn_{i,crit} \simeq 778$. $\omega$ is the frequency and $\gamma$ the growth rate at each point in the domain. The stability boundary is marked with a gray line at $\gamma = 0$. The are seven equally spaced contours from $-1.8$ to 0.}
    \label{fig:pseudospectra}
\end{figure}
\section{Results}
\label{sec:nonlinear}
Figure~\ref{fig:urms_tz_rei700} shows a typical DNS snapshot of the rms velocity perturbation $u'_{rms}$ in the middle of the gap for $\Reyn_i=700$. Despite being well below the linear stability threshold $\Reyn_{i,crit} \simeq 778$, the flow morphology is nearly identical to the supercritical case $\Reyn_i = 900$ (not shown). 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{../figs/urms_tz_rei_700_reo-3398_000278.png}
    \caption{Slice of fluctuating rms velocity in the middle of the radial domain at $\Reyn_i = 700$, $\Reyn_o = -3398$. This solution is well below the linear stability threshold at }
    \label{fig:urms_tz_rei700}
\end{figure}

\subsection{Supercritical Spiral Turbulence}
We first consider the supercritical case with $\Reyn_i = 900$. We initialize GQL simulations with $\Lambda = 0, 1, 3, 5, 10, 30, 60$ using a snapshot of DNS data at $t \simeq 1.5 \tau_\nu$.  Figure~\ref{fig:rei900_snapshots} shows slices at mid-gap, $r = R_{mid}$. The first important thing to note is that low-order models are not able to maintain the spiral structure until $\Lambda \geq 30$. However, even including a single additional low mode, $\Lambda = 1$, shows a significant increase in flow complexity compared to QL. It is quite interesting to note that QL \emph{does} feature some degree of spatial intermittency--there are clearly separated regions hosting spiral waves interspersed with featureless, laminar patches. 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figs/rei900_snapshots.pdf}
    \caption{RMS velocity perturbations in the $z-\theta$ plane at $r=R_{mid}$ for $\Reyn_i=900, \Reyn_0=-3398$. The upper left panel shows DNS, the upper right panel shows QL, and the remaining four are GQL with $\Lambda = 1,3,5,10$. Until $\Lambda > 30$, spiral turbulence is not apparent.}
    \label{fig:rei900_snapshots}
\end{figure}

Figure~\ref{fig:ke_vs_t_rei900} shows the kinetic energy as a function of time and $\Lambda$ for both the supercritical (left) and subcritical (right) runs. 
The supercritical case shows significant oscillations at all $\Lambda < 30$.
While both $\Lambda 30$ and 60 show spiral turbulence, $\Lambda=30$ has considerably more wave motion in the laminar region than either $\Lambda = 60$ or DNS. 
Interestingly, $\Lambda=30$ has a lower kinetic energy much closer to that of DNS than does $\Lambda=60$.
By examining power spectra (figure
\begin{figure}
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/rei900_reo_-3398_KE_vs_t.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/rei700_reo_-3398_KE_vs_t.pdf}
    \end{minipage}
    \caption{Perturbation kinetic energy as a function of time for $\Reyn_o=-3398$. $\Reyn_i = 900$ (left), started from small perturbations, is above the linear stability boundary. $\Reyn_i = 700$ (right panel), continued by lower $\Reyn_i$ from previous simulations, is in the hysteretic region below the stability boundary. Starting from a saturated spiral turbulence state from DNS at $t/\tau_\nu \simeq 1.5$ (left), $t/\tau_\nu \simeq 3$ (right), we continue the solution with QL ($\Lambda = 0$) and GQL with cutoffs $\Lambda = 1,3,5,10,30,60$. At higher $\Reyn_i$, all of the reduced models begin high amplitude oscillations. At lower $\Reyn_i$, $\Lambda = 60$ is required to maintain the spiral turbulence state. }
    \label{fig:ke_vs_t_rei900}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{../figs/rei900_spectra.pdf}
    \caption{$\theta$-$z$ power spectra of $v$ at midgap for $\Reyn_i = 900$ at various GQL cutoffs. Adding a single GQL mode significantly improves the power spectrum. Spiral turbulence appears at $\Lambda=30$. The obvious square in $\Lambda = 60$ represents a sudden drop in spectral power at the cutoff wavenumber.}
    \label{fig:my_label}
\end{figure}
Focusing on $\Lambda=5$, figure~\ref{fig:rei900_lambda5_story} shows that the flow goes through a series of different states bearing a strong resemblance to other well-known TC flow patterns originally identified in exhaustive series of experiments reported in \cite{1984JFM...146...45M} and subsequently found in simulations (e.g. \cite{2009PhRvE..80d6315M}). At the time of kinetic energy minimum, labelled 1 in the left hand side of figure~\ref{fig:rei900_lambda5_story}, the flow seems to recapitulate the interpenetrating spirals (IPS) characteristic of lower $\Reyn$ counter-rotating. As the kinetic energy rises, reaching a knee at point 2, the flow morphology resembles the patchy bursts characteristic of the intermittency regime (INT). Finally, at the peak of the kinetic energy oscillation, point 3 in figure~\ref{fig:rei900_lambda5_story}, the flow reaches its closest approach to something resembling spiral turbulence, though it is very difficult to identify the flow as such from a slice at a single point in time and plane in space.

This is notable for a few reasons. At $\Reyn_i=900, \Reyn_o=-3398$ the system is not in a hysteretic region of parameter space. Thus, we do not expect this behaviour to be due to GQL picking up other, coexisting solutions. 
It also highlights the delicate balance an intermittent laminar/turbulent solution represents. The turbulence in this solution has a very broad range of scales, as evidenced by figure~\ref{fig:urms_tz_rei700}. Understanding how it exchanges energy with the mean flow is crucial to understanding both the saturation of the underlying linear instability as well as the maintenance of the non-linear state. 
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/rei900_story.pdf}
    \caption{Kinetic energy (left) and flow morphologies (right) for $\Reyn_i = 900$, $\Lambda = 5$ covering one full oscillation cycle. The panels on the right show three times corresponding to the vertical lines in the kinetic energy plot on the left. At the kinetic energy minimum (point 1), the flow shows a IPS-like morphology, while an INT burst occurs at the knee (point 2), and a more space-filling turbulent state occurs at maximum energy (point 3).}
    \label{fig:rei900_lambda5_story}
\end{figure}
Finally, in figure~\ref{fig:rei900_vmean}, we show the azimuthal velocity averaged over $\theta$ and $z$, giving the mean deviation from CCF as a function of GQL cutoff $\Lambda$. Note that no conditional averaging \cite{2009PhRvE..80f7301D} is performed here, so this represents an average over both laminar and turbulent regions for DNS. As $\Lambda$ is increased, the profile approaches the DNS case; this is typical of GQL behaviour (figure 7 of \cite{2017JFM...810..412T} shows the same for rotating plane Couette flow). However, QL performs slightly better than low order GQL with $\Lambda < 5$. 
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/rei900_reo_-3398_vmean_profile.pdf}
    \caption{Mean deviation from CCF for $\Reyn_i = 900$, $\Reyn_o=-3398$. The heavy line is DNS, the orange line is QL, and all others are GQL with $\Lambda = 1,3,5,10$.}
    \label{fig:rei900_vmean}
\end{figure}

@@

\section{Conclusion}
\label{sec:conclusion}
We have presented a series of simulations demonstrating the performance of GQL at various cutoffs for Taylor-Couette flow in the spiral turbulence regime. GQL is a significant improvement over QL models in representing statistical quantities, particularly the power spectra. However, it does not retain an unambiguous spiral form in the supercritical regime until $\Lambda \simeq 30$; at subcritical $\Reyn_i$, $\Lambda \simeq 60$ is required to see any sustained flows at all. Interestingly, when GQL solutions do sustain non-CCF flows in the subcritical regime, they also produce spiral patterns. Prior work has focused on the utility of GQL in presenting a more accurate \emph{statistical} picture of turbulent flows than QL \cite{2016PhRvL.116u4501M, 2017JFM...810..412T,2018RSPSA.47480422T,2019Kellam}. This focus is particularly important in light of its potential as as means of improving the accuracy of direct statistical simulation (DSS), a method for studying turbulence by computing its statistics directly, rather than individual realizations \cite{2022arXiv220505513M}. 
We have shown here that it is possible to use GQL as a means to investigate the non-linear interactions of greatest importance for maintaining the intricate patterns manifested in Taylor-Couette flow. In future work, we will consider the ability of GQL to maintain exact coherent states identified by Deguchi et al \cite{2014PhRvL.112r4502D} and the dynamics of which were studied recently by Wang and coworkers \cite{2022arXiv220712990W}. Understanding how these interactions work to shape subcritical TC turbulence represents a promising new avenue to understanding how ECS are maintained and how they in turn sustain turbulence below linear stability bounds. In particular, one can begin to understand the relative importance of non-normal effects (represented entirely with GQL) and non-linear ECS.
